A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-15 text/formal-defnition.tex
Add examples.
Fix by grammarly.

% \AI{Rename ``type variables'' to ``type-level constants''. }
% \AI{I prefer $\varepsilon$ for the nonempty sequence.}  
We assume the denumerable set of type-level constants, ranged over by
metavariables \(X, Y, Z\), the denumerable set of variables, ranged
over by \(x,y,z\), the denumerable set of constants, ranged over by
\(c\), and the denumerable set of stage variables, ranged over by
\(\alpha, \beta, \gamma\).  The metavariables \(A, B, C\) range over
sequences of stage variables; we write \(\varepsilon\) for the empty
sequence. \LMD is defined by the following grammar:

@@

% Description on meta variables

@@

We adopt a \LLF-like system, where constants and type-level constants are given in the signature $\Sigma$, which
are a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use integers in \LMD, $\Sigma = \B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Please note that constants or type-level constants has no stage.
This means that we can use them at any stage.
% Type environments
Type environments are sequences of triples of a variable, its type, and its stage.
We define well-formed signature and well-formed type environments later.

@@

As the end of this subsection, we compare \LMD and \LTP.
Terms of \LMD is the same with \LTP except constants.
The difference between \LMD and \LTP is their types and kinds.
First, \LMD contains $\Pi$ type instead of function type in \LTP.
Second, there two kinds of $\F$ in \LTP in order to disscuss on program residualization
although \LMD contains only one $\F$ type.
Third, \LMD contains kinding and type equivalence rules to handle dependent types.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are 
the ordinary capture-avoiding substitutions of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively
and we omit their definition here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are 
substitutions of stage $A$ for stage variable $\alpha$ in 
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$ respectively.
We show replesentive cases below.
% \AI{What about $\tau[x \mapsto N], K[x \mapsto N], \tau [\alpha \mapsto \varepsilon]$ and $K[\alpha \mapsto \varepsilon]$?}

@@

The relation $\longrightarrow_\beta$ represents ordinary $\beta-$reduction in the \(\lambda\)-calculus; the relation
$\longrightarrow_\blacklozenge$ represents that quotation $\TB_\alpha M$ is canceled by escape and $M$ is spliced into the code fragment surrounding the escape;
the relation $\longrightarrow_\Lambda$ means that a stage abstraction applied to the empty stage $\varepsilon$ reduces to the body of abstraction
where $\varepsilon$ is substituted for the stage variable.
There is no reduction rule for CSP as with Hanada and Igarashi \cite{Hanada2014}.
The CSP operator $\%_\alpha$ disappears when $\varepsilon$ is substituted to $\alpha$.
We show an example of reduction sequence below.
Underlines show the redexes.
\begin{align*}
	 & \hspace{10mm} \underline{(\lambda i:\I\to\I.(\Lambda\alpha.\TB_\alpha (\%_\alpha i\ 1 + (\TBL_\alpha \TB_\alpha 3))\ \varepsilon))\ (\lambda x:\I.x)} \\
	 & \longrightarrow_\blacklozenge (\Lambda\alpha.\TB_\alpha (\%_\alpha (\lambda x:\I.x)\ 1 + (\underline{\TBL_\alpha \TB_\alpha 3})))\ \varepsilon        \\
	 & \longrightarrow_\beta \underline{(\Lambda\alpha.\TB_\alpha (\%_\alpha (\lambda x:\I.x)\ 1 + 3))\ \varepsilon}                                         \\
	 & \longrightarrow_\Lambda \underline{(\lambda x:\I.x)\ 1} + 3                                                                                           \\
	 & \longrightarrow_\beta 1 + 3                                                                                                                           \\
	 & \longrightarrow^* 4
\end{align*}

@@

% \AI{Insert a tilde (which is a space without a line break) between ``Figure'' and ref.}
\begin{definition}[Typing, Kinding, Well-formed Kinding, Term Equivalence, Type Equivalence, Kind Equivalence]
	The typing relation $ \G \V M : \tau @ A $,
	the kinding relation $\G \V \tau :: K @ A$,
	the well-formed kinding relation $\G \V K \iskind @ A$,
	the term equivalence relation $\G \V M \E N @ A$,
	the type equivalence relation $\G \V \tau \E \sigma @ A$,
	the kind equivalence relation $\G \V K \E J @ A$,
	are the least relation closed under the rules in Figure~\ref{fig:typing-rules}.
\end{definition}
This definition is little complicated to define six relations at the same time.
We canot define them one by one because they are dependent on each other.

@@

% \TConv
\AI{Looks like a paper is written for those who know multi-stage calculi fairly well but don't know dependent type systems.  The reality is opposite.}
As in standard dependent type systems, \TConv{} allows us to replace
the type of a term with an equivalent one.
In a type system which includes dependent types, this kind of rule is essential
because two types which have different shapes may be equivalent.
For example, when we use a vector type with its size ($\textrm{Vector}\ n$),
$\textrm{Vector}\ 5$ is equivalent to $\textrm{Vector}\ (4+1)$ obviously although they are not equivalent in apparently.

% Typing rules for a multi-stage calculus
% \AI{This paragraph is not very informaticve.  We shouldn't just refer redears to previous work.}
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are rules for a multi-stage calculus.
The rule \TTB{}, which corresponds to brackets, means that if term $M$ is of type $\tau$ at stage $A\alpha$,
$\TB_\alpha M$ is of type $\TW_\alpha \tau$ at stage $A$.
The rule \TTBL{}, which corresponds to escape, is the converse of \TTB.
The rule \TGen{} is a rule for stage abstraction.
The condition of $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensure that
substitutions to $\alpha$ don't affect type environments or stages.
The rule \TIns{} is for applications of stages to stage abstractions.
The rule \TCsp is the rule for CSP, which means that 
if term $M$ is of type $\tau$ at stage $A$, $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.

@@

Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction to remove $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator
in implementation.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense and as we argue next
it is practically significant.

@@

The reduction given above is full reduction and any redexes---even
under quotation---can be reduced in an arbitrary order.  
Following previous work~\cite{Hanada2014}, 
we introduce (small-step, call-by-value) staged semantics, 
where only $\beta$-reduction or $\Lambda$-reduction at stage or the outer-most $\blacklozenge$-reduction are allowed, 
modeling an implementation.

@@

\begin{definition}[Values]
	The family $V^A$ of sets of values, ranged over by $v^A$,
	is defined by following grammar.  In the grammar, $A \neq \varepsilon$ is assumed.
	\begin{align*}
		v^\varepsilon \in V^\varepsilon & ::= \lambda x:\tau.M \mid\ \TB_\alpha v^\alpha \mid \Lambda\alpha.v^\varepsilon                                       & \\
		v^A \in V^A                     & ::= x \mid \lambda x:\tau.v^A \mid v^A\ v^A \mid\ \TB_\alpha v^{A\alpha} \mid \Lambda\alpha.v^A \mid v^A\ \varepsilon & \\
		                                & \quad\   \mid\ \TBL_\alpha v^{A'} (\text{if } A = A'\alpha \text{ for some } \alpha, A' \neq \varepsilon)             & \\
		                                & \quad\   \mid\ \%_\alpha v^{A'} (\text{if } A = A'\alpha  \text{ for some } \alpha, A')
	\end{align*}
\end{definition}

@@

In order to define Staged Reduction, we define Redex and Evaluation Context.

@@

We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$ 
and define the staged reduction relation.
Intuitively, $\square$ is the placefolder for redex.
Examples of evaluation context is below.
\red{Example of evaluation context}
% \begin{align*}
% \end{align*}

@@

This reduction relation reduces a typed term in a unique way 
which is appropriate as call-by-value code generater.
It means that terms in brackets is not evaluated until the terms are run
and arguments of a function are evaluated before the application.
\red{Example of staged reduction}

>>>>

% \AI{Rename ``type variables'' to ``type-level constants''. }
% \AI{I prefer $\varepsilon$ for the non-empty sequence.}  
We assume the denumerable set of type-level constants, ranged over by
metavariables \(X, Y, Z\), the denumerable set of variables, ranged
over by \(x,y,z\), the denumerable set of constants, ranged over by
\(c\), and the denumerable set of stage variables, ranged over by
\(\alpha, \beta, \gamma\).  The metavariables \(A, B, C\) range over
sequences of stage variables; we write \(\varepsilon\) for the empty
sequence. \LMD is defined by the following grammar:

@@

% Description of meta variables

@@

We adopt a \LLF-like system, where constants and type-level constants are given in the signature $\Sigma$, which
are a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use integers in \LMD, $\Sigma = \B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Please note that constants or type-level constants have no stage.
This means that we can use them at any stage.
% Type environments
Type environments are sequences of triples of a variable, its type, and its stage.
We define well-formed signature and well-formed type environments later.

@@

As the end of this subsection, we compare \LMD and \LTP.
Terms of \LMD are the same with \LTP except for constants.
The difference between \LMD and \LTP is their types and kinds.
First, \LMD contains $\Pi$ type instead of function type in \LTP.
Second, there two kinds of $\F$ in \LTP in order to discuss on program residualization
although \LMD contains only one $\F$ type.
Third, \LMD contains kinding and type equivalence rules to handle dependent types.

@@

Next, we define the full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
the ordinary capture-avoiding substitutions of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively
and we omit their definition here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$ respectively.
We show replesentive cases below.
% \AI{What about $\tau[x \mapsto N], K[x \mapsto N], \tau [\alpha \mapsto \varepsilon]$ and $K[\alpha \mapsto \varepsilon]$?}

@@

The relation $\longrightarrow_\beta$ represents ordinary $\beta-$reduction in the \(\lambda\)-calculus; the relation
$\longrightarrow_\blacklozenge$ represents that quotation $\TB_\alpha M$ is canceled by escape and $M$ is spliced into the code fragment surrounding the escape;
the relation $\longrightarrow_\Lambda$ means that a stage abstraction applied to the empty stage $\varepsilon$ reduces to the body of the abstraction
where $\varepsilon$ is substituted for the stage variable.
There is no reduction rule for CSP as with Hanada and Igarashi \cite{Hanada2014}.
The CSP operator $\%_\alpha$ disappears when $\varepsilon$ is substituted to $\alpha$.
We show an example of a reduction sequence below.
Underlines show the redexes.
\begin{align*}
     & \hspace{10mm} \underline{(\lambda i:\I\to\I.(\Lambda\alpha.\TB_\alpha (\%_\alpha i\ 1 + (\TBL_\alpha \TB_\alpha 3))\ \varepsilon))\ (\lambda x:\I.x)} \\
     & \longrightarrow_\blacklozenge (\Lambda\alpha.\TB_\alpha (\%_\alpha (\lambda x:\I.x)\ 1 + (\underline{\TBL_\alpha \TB_\alpha 3})))\ \varepsilon        \\
     & \longrightarrow_\beta \underline{(\Lambda\alpha.\TB_\alpha (\%_\alpha (\lambda x:\I.x)\ 1 + 3))\ \varepsilon}                                         \\
     & \longrightarrow_\Lambda \underline{(\lambda x:\I.x)\ 1} + 3                                                                                           \\
     & \longrightarrow_\beta 1 + 3                                                                                                                           \\
     & \longrightarrow^* 4
\end{align*}

@@

% \AI{Insert a tilde (which is a space without a line break) between ``Figure'' and ref.}
\begin{definition}[Typing, Kinding, Well-formed Kinding, Term Equivalence, Type Equivalence, Kind Equivalence]
    The typing relation $ \G \V M : \tau @ A $,
    the kinding relation $\G \V \tau :: K @ A$,
    the well-formed kinding relation $\G \V K \iskind @ A$,
    the term equivalence relation $\G \V M \E N @ A$,
    the type equivalence relation $\G \V \tau \E \sigma @ A$,
    the kind equivalence relation $\G \V K \E J @ A$,
    are the least relation closed under the rules in Figure~\ref{fig:typing-rules}.
\end{definition}
This definition is little complicated to define six relations at the same time.
We can't define them one by one because they are dependent on each other.

@@

% \TConv
\AI{Looks like a paper is written for those who know multi-stage calculi fairly well but don't know dependent type systems.  The reality is the opposite.}
As in standard dependent type systems, \TConv{} allows us to replace
the type of a term with an equivalent one.
In a type system which includes dependent types, this kind of rule is essential
because two types which have different shapes may be equivalent.
For example, when we use a vector type with its size ($\textrm{Vector}\ n$),
$\textrm{Vector}\ 5$ is equivalent to $\textrm{Vector}\ (4+1)$ obviously although they are not equivalent in apparently.

% Typing rules for a multi-stage calculus
% \AI{This paragraph is not very informative.  We shouldn't just refer readers to previous work.}
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are rules for a multi-stage calculus.
The rule \TTB{}, which corresponds to brackets, means that if term $M$ is of type $\tau$ at stage $A\alpha$,
$\TB_\alpha M$ is of type $\TW_\alpha \tau$ at stage $A$.
The rule \TTBL{}, which corresponds to escape, is the converse of \TTB.
The rule \TGen{} is a rule for stage abstraction.
The condition of $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensure that
substitutions to $\alpha$ don't affect type environments or stages.
The rule \TIns{} is for applications of stages to stage abstractions.
The rule \TCsp is the rule for CSP, which means that
if term $M$ is of type $\tau$ at stage $A$, $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.

@@

Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction to remove $\%_\alpha$ when they developed \LTP{}, as such
the reduction does not match the operational behavior of the CSP operator
in implementation.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense and as we argue next
it is practically significant.

@@

The reduction given above is full reduction and any redexes---even
under quotation---can be reduced in an arbitrary order.
Following previous work~\cite{Hanada2014},
we introduce (small-step, call-by-value) staged semantics,
where only $\beta$-reduction or $\Lambda$-reduction at stage $\varepsilon$ or the outer-most $\blacklozenge$-reduction are allowed,
modeling an implementation.

@@

\begin{definition}[Values]
    The family $V^A$ of sets of values, ranged over by $v^A$,
    is defined by the following grammar.  In the grammar, $A \neq \varepsilon$ is assumed.
    \begin{align*}
        v^\varepsilon \in V^\varepsilon & ::= \lambda x:\tau.M \mid\ \TB_\alpha v^\alpha \mid \Lambda\alpha.v^\varepsilon                                       & \\
        v^A \in V^A                     & ::= x \mid \lambda x:\tau.v^A \mid v^A\ v^A \mid\ \TB_\alpha v^{A\alpha} \mid \Lambda\alpha.v^A \mid v^A\ \varepsilon & \\
                                        & \quad\   \mid\ \TBL_\alpha v^{A'} (\text{if } A = A'\alpha \text{ for some } \alpha, A' \neq \varepsilon)             & \\
                                        & \quad\   \mid\ \%_\alpha v^{A'} (\text{if } A = A'\alpha  \text{ for some } \alpha, A')
    \end{align*}
\end{definition}

@@

In order to define Staged Reduction, we define redex and evaluation context.

@@

We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$
and define the staged reduction relation.
Intuitively, $\square$ is the placefolder for redex.
Examples of evaluation context is below.
\begin{align*}
     & \square\ (\lambda x:\I.x) \in  ECtx^\varepsilon_\varepsilon            \\
     & \Lambda\alpha.\square\ \epsilon \in ECtx^\varepsilon_\varepsilon       \\
     & \TBL_\alpha \TB_\alpha \TBL_\alpha \square \in ECtx^\alpha_\varepsilon
\end{align*}

@@

This reduction relation reduces a typed term in a unique way
which is appropriate as call-by-value code generator.
It means that terms in brackets are not evaluated until the terms are run
and arguments of a function are evaluated before the application.
We show an example of staged reduction.
Underlines show the redexes.
\begin{align*}
     & \hspace{9mm} \Lambda\alpha.(\TB_\alpha \underline{\TBL_\alpha \TB_\alpha} (\lambda x:\I.x)\ 10)\ \varepsilon \\
     & \longrightarrow_s \underline{\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x)\ 10)\ \varepsilon}                   \\
     & \longrightarrow_s \underline{(\lambda x:\I.x)\ 10}                                                           \\
     & \longrightarrow_s 10
\end{align*}
